Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is a merged and final version of two conference papers, \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}, with a unified and streamlined presentation.}

### 2020-07-15 journal paper/itcs18paper/app-examples.tex
Lots of edits and comments in the "theory" appendix.

\newcommand{\ExplorExploit}{\term{ExplorExploit}}
\newcommand{\PhasedExplorExploit}{\term{PhasedExplorExploit}}
\newcommand{\RandomDynGreedy}{\term{RandomDynGreedy}}
\newcommand{\SuccesiveElimination}{\term{SuccesiveElimination}}
\newcommand{\SuccesiveEliminationReset}{\term{SuccesiveEliminationReset}}

This appendix provides some pertinent background on multi-armed
bandits (\emph{MAB}). We discuss \BIR and monotonicity of several MAB algorithms, touching upon: \DynGreedy and \StaticGreedy (Section~\ref{sec:MAB-greedy}), ``naive" MAB algorithms that separate exploration and exploitation (Section~\ref{sec:MAB-naive}), and ``smart" MAB algorithms that combine exploration and exploitation (Section~\ref{sec:MAB-smart}).

As we do throughout the paper, we focus on MAB with i.i.d. rewards and a Bayesian prior; we call it \emph{Bayesian MAB} for brevity.

@@

\subsection{\DynGreedy and \StaticGreedy}
\label{sec:MAB-greedy}

We provide an example when \DynGreedy and \StaticGreedy have
constant \BIR, and prove monotonicity of \DynGreedy. For the
example, it suffices to consider \emph{deterministic rewards} (for
each action $a$, the realized reward is always equal to the mean
$\mu_a$) and \emph{independent priors} (according to the prior
$\priorMu$, random variables $\mu_1 \LDOTS \mu_K$ are mutually
independent) each of {\em full support}.

%\ascomment{This lemma is just one example. Can we argue that constant \BIR is typical/reasonable?}
%
%
%\begin{lemma}
%There is a problem instance of Bayesian MAB such that \StaticGreedy
%and \DynGreedy have (at least) a constant \BIR for all steps. The
%problem instance has two actions, independent priors, and
%deterministic rewards.
%\end{lemma}
%
%\begin{proof}
%To specify the problem instance, it remains to specify the marginal
%distributions mean rewards $\mu_1$ and $\mu_2$ according to the
%prior. We posit that $\mu_1$ is uniform on the interval $[0, 1]$,
%and $\mu_2$ is uniform on the interval $[\tfrac13,1]$.
%
%Note that
%    $\OPT := \E[\max_{a\in A} \mu_a] =(1/3)(2/3)+(2/3)(7/9)=20/27$.
%%\ascomment{The numbers seem wrong, given the below?}
%This holds since with
%probability $1/3$ we have $\mu_1\leq 1/3$ in which case clearly the
%better arm is 2 with $\E[\mu_2]=2/3$ and with probability $2/3$ we
%have both $\mu_1,\mu_2\in[1/3,1]$ and distributed uniformly each.
%
%\StaticGreedy always recommends action $2$. Since
%$\E[\mu_2]=\tfrac23$, \BIR is $2/27$ in all rounds.
%
%\DynGreedy recommends the first agent action $2$. If $\mu_2 > 2/3$
%For all the following agents it recommends action $2$. Otherwise, it
%recommends agent $2$ action $1$ and for all the following agents it
%recommends the better action. The expectation for agents $t\geq 3$
%is $(1/2)(5/6)+(1/2)((1/2)(1/2)+(1/2)(5/9))=49/72$ and has
%instantaneous regret $13/216$.
%\end{proof}

The following claim is immediate from the definition of the CDF
function
\begin{claim}
Assume independent priors. Let $F_i$ be the CDF of the mean reward
$\mu_i$ of action $a_i\in A$. Then, for any numbers
$z_2>z_1>\E[\mu_2]$ we have
    $\Pr[\text{$\mu_1\leq z_1$ and $\mu_2\geq z_2$}] = F_1(z_1)(1-F_2(z_2)) $.
\end{claim}

%\begin{lemma}
%Let $\priorMu$ be the prior of the means, and assume that it is
%independent for each action, where for arm $i$ it has a mean $
%\E[\mu_i]$ and CDF $F_i$. Then, for any $\E[\mu_2] < z_1<z_2$ we
%have that with probability $F_1(z_1)(1-F_2(z_2))$ that both
%$\mu_1\leq z_1$ and $z_2\leq \mu_2$.
%\end{lemma}

%\begin{corollary}
%Any problem instance of Bayesian MAB for two actions and independent
%priors which are full support and deterministic rewards, with
%constant probability \DynGreedy has a constant \BIR for all steps.
%\end{corollary}

We can now draw an immediate corollary of the above claim

\begin{corollary}
Consider any problem instance of Bayesian MAB with two actions and independent
priors which are full support. Then:
\begin{OneLiners}
\item[(a)] With constant probability, \StaticGreedy  has a constant \BIR for all steps.
\item[(b)] Assuming deterministic rewards, with
constant probability \DynGreedy has a constant \BIR for all steps.
\end{OneLiners}
\end{corollary}

\begin{remark}
A similar result holds for  rewards which are distributed as
Bernoulli random variables. In this case we consider accumulative
reward of an action as a random walk, and use a high probability
variation of the law of iterated logarithms. (Details omitted.)
\end{remark}

Next, we show that \DynGreedy is monotone.

\begin{lemma}\label{dgmono}
\DynGreedy is monotone, in the sense that $\rew(n)$ is non-decreasing.
Further, $\rew(n)$ is strictly increasing for every time step $n$ with $\Pr[a_n\neq a_{n+1}]>0$.
\end{lemma}

@@

\subsection{``Naive" MAB algorithms that separate exploration and exploitation}
\label{sec:MAB-naive}

@@

MAB algorithm \ExplorExploit$(m)$ initially explores each action
with $m$ agents and for the remaining $T-|A|m$ agents recommends the
action with the highest observed average. In the explore phase it
assigns a random permutation of the $mK$ recommendations.

\begin{lemma}
%Assume that rewards in the range $[0,1]$.
The \ExplorExploit$(T^{2/3}\log |A|/\delta)$ algorithm has, with
probability $1-\delta$, for any  $n\geq |A|T^{2/3}$ we have
\BIR$(n)=O(T^{-1/3})$. In addition, \ExplorExploit$(m)$ is monotone.
\end{lemma}

@@

We can also have a a phased version which we call
\PhasedExplorExploit$(m_t)$, where time is partition in to phases.
In phase $t$ we have $m_t$ agents and a random subset of $K$ explore
the actions (each action explored by a single agent) and the other
agents exploit. (This implies that we need that $m_t\geq K$ for all
$t$. We also assume that $m_t$ is monotone in $t$.)

\begin{lemma}
Consider the case that $K=2$ and the rewards of the actions are
Bernoulli r.v. with parameter $\mu_i$ and $\Delta=\mu_1-\mu_2$.
Algorithm \PhasedExplorExploit$(m_t)$ is monotone and for $m_t =
\sqrt{t}$ it has $\BIR(n)=O(n^{-1/3}+e^{-O(\Delta^2 n^{2/3})}))$. 
\end{lemma}

\begin{proof}
We first show that it is monotone. Recall that $\mu_1>\mu_2$. Let
$S_i=\sum_{j=1}^t r_{i,j}$ be the sum of the rewards of action $i$
up to phase $t$. We need to show that $\Pr[S_1>S_2]+ (1/2)
\Pr[S_1=S_2]$ is monotonically increasing in $t$. Consider the
random variable $Z=S_1-S_2$. At each phase it increases by $+1$ with
probability $\mu_1(1-\mu_2)$, decreases by $-1$ with probability
$(1-\mu_1)\mu_2$ and otherwise does not change.

@@

\subsection{``Smart" MAB algorithms that combine exploration and exploitation}
\label{sec:MAB-smart}

%MAB algorithm \SuccesiveElimination holds a set of action
%$A_s\subset A$ which are called {\em surviving actions} and for
%agent $t$ selects a random action from $A_s$. Let $n_{i,t}$ be the
%number of times action $i$ has been selected up to time $t$, and
%$\hat{\mu}_{i,t}$ be the average of the rewards of action $i$ up to
%time $t$ and $\hat{\mu}^*=\max_i \hat{\mu}_{i,t}$. We eliminate
%action $i$ at time $t$, i.e., delete it from $A_s$, if
%$\hat{\mu}_t^*-\hat{\mu}_{i,t} > \log(T/\delta)/\sqrt{n/K}$.
%
%\begin{lemma}
%Assume that the prior is independent and if $\mu_i\geq \mu_j$ then
%$r_i$ stochastically dominates $r_j$.
%%
%The algorithm \SuccesiveElimination, has, with probability
%$1-\delta$, \BIR$(n)=O(\log(T/\delta)/\sqrt{n/K})$.
%\end{lemma}

@@

MAB algorithm \SuccesiveEliminationReset works as follows. It keeps
a set of surviving actions $A_s\subseteq A$, where initially
$A_s=A$. The agents are partition into phases, where each phase is a
random permutation of the non-eliminated actions.
% Initially, we select actions at random. Let $n_{i,t}$
%be the number of times action $i$ has been selected up to time $t$,
Let $\hat{\mu}_{i,t}$ be the average of the rewards of action $i$ up
to phase $t$ and $\hat{\mu}^*=\max_i \hat{\mu}_{i,t}$. We eliminate
action $i$ at the end of phase $t$, i.e., delete it from $A_s$, if
$\hat{\mu}_t^*-\hat{\mu}_{i,t} > \log(T/\delta)/\sqrt{t}$.
%So far identical to \SuccesiveElimination, the difference is how we
%continue after elimination.
In \SuccesiveEliminationReset we simply reset the algorithm with
$A=A_s-A_{e,t}$, where $A_{e,t}$ is the set of eliminated actions
after phase $t$. Namely, we restart $\hat{\mu}_{i,t}$ and ignore the
old rewards before the elimination.

\begin{lemma}
%
The algorithm \SuccesiveEliminationReset, has, with probability
$1-\delta$, \BIR$(n)=O(\log(T/\delta)/\sqrt{n/K})$.
\end{lemma}

\begin{proof}
Let the best action be $a^*=\arg\max_a \mu_a$. With probability
$1-\delta$ at any time $n$ we have that for any action $i\in A_s$
that $|\hat{\mu}_i -\mu_i|\leq \log(T/\delta)/\sqrt{n/K}$, and
$a^*\in A_s$. This implies that any action $a$ such
$\mu_{a^*}-\mu_{a}> 3\log(T/\delta)/\sqrt{n/K}$ is eliminated.
Therefore, any action in $A_s$ has \BIR$(n)$ of at most
$6\log(T/\delta)/\sqrt{n/K}$.
\end{proof}

\begin{lemma}
Assume that if $\mu_i\geq \mu_j$ then the rewards $r_i$
stochastically dominates the rewards $r_j$.
%
Then, \SuccesiveEliminationReset is monotone
\end{lemma}

\begin{proof}
Consider the first time $T$ an action is eliminated, and let
$T=\tau$ be a realized value of $T$. Then, clearly for $n<\tau$ we
have $\rew(n)=\rew(1)$ .

>>>>

\newcommand{\ExplorExploit}{\term{ExplorExploit}}
\newcommand{\PhasedExplorExploit}{\term{PhasedExplorExploit}}
%\newcommand{\RandomDynGreedy}{\term{RandomDynGreedy}}
%\newcommand{\SuccesiveElimination}{\term{SuccesiveElimination}}
\newcommand{\SuccesiveEliminationReset}{\term{SuccesiveEliminationReset}}

\newcommand{\IReg}{R^{\term{inst}}} % instantaneous regret

%This appendix provides some pertinent background on multi-armed
%bandits (\emph{MAB}). We discuss \BIR and monotonicity of several MAB algorithms, touching upon: \DynGreedy and \StaticGreedy (Section~\ref{sec:MAB-greedy}), ``naive" MAB algorithms that separate exploration and exploitation (Section~\ref{sec:MAB-naive}), and ``smart" MAB algorithms that combine exploration and exploitation (Section~\ref{sec:MAB-smart}). As we do throughout the paper, we focus on MAB with i.i.d. rewards and a Bayesian prior; we call it \emph{Bayesian MAB} for brevity.

@@

\ascomment{TODO: insert an intro para.}

Throughout this section, we consider a standard MAB setting. There are $T$ rounds and $K$ arms. In each round $t\in [T]$, the algorithm chooses an arm $a_t\in A$ and receives a reward $r_t\in[0,1]$ for this arm, drawn from a fixed but unknown distribution. The set of all arms is $A$; mean reward of arm $a$ is denoted $\mu_a$.

Throughout, we let
    $\rew(t) = \mu_{a_t}$
denote the instantaneous mean reward of the algorithm.

Instantaneous regret is denoted
    $\IReg(t) := \max_{a\in A} \mu_a - \mu_{a_t}$,
so that 
    $\BIR(t) = \E[\IReg(t)]$.

\subsection{Monotonicity-in-information for the greedy step}
\label{sec:MAB-greedy}

\ascomment{TODO: insert an intro sentence.}

\ascomment{Steven: NB: we need monotonicity-in-info to argue about "a little greedy goes a long way".}

The $n$-step history is the random sequence 
    $H_n = \rbr{ (a_t,r_t):\, t\in[n]}$.
Realizations of $H_n$ are called \emph{realized histories}. 
Let $\mH_n$ be the set of all possible values of $H_n$.

The \emph{Bayesian-greedy step} given an $n$-step history $h\in \mH_n$ is defined as
\[ \DG(h) := \argmax_{a\in A} \E\sbr{ \mu_a\mid H_n = h },
\quad\text{ties broken uniformly}.\]

\begin{lemma}\label{lm:MII}
Let $h,h'$ be two realized histories such that $h$ is a prefix of $h'$. Then
\[ \E\sbr{ \mu_{\DG(h)} } \leq \E\sbr{ \mu_{\DG(h')} }. \]
\end{lemma}
\begin{corollary}
\DynGreedy is monotone.
\end{corollary}

\ascomment{TODO: find a citation for Lemma~\ref{lm:MII}.}

\ascomment{OLD TEXT:}

\begin{lemma}\label{dgmono}
\DynGreedy is monotone, in the sense that $\E[\rew(n)]$ is non-decreasing.
\ascomment{Steven: do we ever need the second sentence here?}
Further, $\E[\rew(n)]$ is strictly increasing for every time step $n$ with $\Pr[a_n\neq a_{n+1}]>0$.
\end{lemma}

@@

\subsection{Monotone MAB algorithms}
\label{app:MAB-monotone}

\ascomment{Steven: NB: this is the point of this subsection...}
This subsection argues that ``monotonicity" is a mild assumption in our context. To this end, we present versions of some standard MAB algorithms that are monotone and satisfy ``standard" regret bounds.

\subsubsection{Monotone, exploration-separating MAB algorithms}
\label{app:MAB-monotone-naive}

@@

We provide two exploration-separating algorithms that are monotone and achieve $\IReg(t)=\tildeO\rbr{t^{-1/3}}$. (As we saw in Appendix~\ref{sec:related-classes}, this is the ``right" regret rate for exploration-separating algorithms.)

\ascomment{Steven: can't we prove that Epsilon-Greedy is monotone, too?}

We start with a simple \emph{Explore-First} algorithm:

\begin{lemma}
Consider an algorithm which initially explores each action $m$ times (where $m$ is a parameter), chooses an action  with the highest observed average reward, and uses it for the remaining steps. In the initial exploration phase, actions are randomly permuted. 
\begin{itemize}
\item[(a)] This algorithm is monotone, for every \MRV.
\ascomment{Steven: right?}

\item[(b)] Fix $\delta>0$ and 
    $m = T^{2/3}\log(|A|/\delta)$.
Then, with probability $1-\delta$, it holds that  
    $\IReg(t)=\tildeO\rbr{T^{-1/3}}$
for each step $n> m\,|A|$.
\ascomment{Steven: can you derive a statement on regret, too?}
\end{itemize}
\end{lemma}

@@

We consider a slightly more advanced version, called \PhasedExplorExploit, which enjoys a similar regret bound uniformly over all steps.

The algorithm is defined as follows. The time is partition in to phases $i = 1,2, \ldots $ of $m_i\geq K$ rounds each, where $m_i$ is chosen in advance and increase with $i$. In each phase $i$, $K$ rounds are chosen at random for exploration, and are assigned randomly to one arm each. The remaining rounds are for exploitation. \ascomment{Steven: exploit how? Bayesian or frequentist? Do we use the data from the same phase?}

\begin{lemma}
Posit $K=2$ arms and Bernoulli rewards. 
\begin{itemize}
\item[(a)] Algorithm \PhasedExplorExploit is monotone, for every \MRV.
    \ascomment{Steven: right?}

\item[(b)] For phase durations $m_i =\sqrt{i}$, it has 
    $\IReg(n)=O(n^{-1/3}+e^{-O(\Delta^2 n^{2/3})}))$,
where $\Delta = |\mu_1-\mu_2|$.
\ascomment{Steven: can you derive $O(T^{2/3})$ regret, too?}
\end{itemize}
\end{lemma}

\begin{proof}
We first show that it is monotone. Recall that $\mu_1>\mu_2$. 
\ascomment{Steven: we only have it for prior means, \ie 
$\E[\mu_1]>\E[\mu_2]$}

Let
$S_i=\sum_{j=1}^t r_{i,j}$ be the sum of the rewards of action $i$
up to phase $t$. We need to show that $\Pr[S_1>S_2]+ (1/2)
\Pr[S_1=S_2]$ is monotonically increasing in $t$. Consider the
random variable $Z=S_1-S_2$. At each phase it increases by $+1$ with
probability $\mu_1(1-\mu_2)$, decreases by $-1$ with probability
$(1-\mu_1)\mu_2$ and otherwise does not change.

@@

\subsubsection{Monotone, adaptive-exploration MAB algorithms}
\label{app:MAB-monotone-smart}

We provide a simple modification of Successive Elimination algorithm from \citep{EvenDar-icml06} that is monotone and (still) enjoys the same bounds on regret and BIR. 
%MAB algorithm \SuccesiveElimination holds a set of action
%$A_s\subset A$ which are called {\em surviving actions} and for
%agent $t$ selects a random action from $A_s$. Let $n_{i,t}$ be the
%number of times action $i$ has been selected up to time $t$, and
%$\hat{\mu}_{i,t}$ be the average of the rewards of action $i$ up to
%time $t$ and $\hat{\mu}^*=\max_i \hat{\mu}_{i,t}$. We eliminate
%action $i$ at time $t$, i.e., delete it from $A_s$, if
%$\hat{\mu}_t^*-\hat{\mu}_{i,t} > \log(T/\delta)/\sqrt{n/K}$.
%
%\begin{lemma}
%Assume that the prior is independent and if $\mu_i\geq \mu_j$ then
%$r_i$ stochastically dominates $r_j$.
%%
%The algorithm \SuccesiveElimination, has, with probability
%$1-\delta$, \BIR$(n)=O(\log(T/\delta)/\sqrt{n/K})$.
%\end{lemma}

@@

The algorithm, called \SuccesiveEliminationReset, proceeds in phases. In each phase $j$, the algorithm is given a set of ``surviving" arms $A_j\subseteq A$, with $A_1=A$. The phase lasts $|A_j|$ rounds, randomly permuted across the arms in $A_j$ (so that each arm in $A_j$ is chosen exactly once).

\ascomment{Stopped here.}

\ascomment{Steven: I don't understand the algorithm! It resets after each phase, right? But then, the history is just one sample per each arm? WTF?}

Let $\hat{\mu}_{i,t}$ be the average reward of action $i$ up
to phase $t$ and $\hat{\mu}^*=\max_i \hat{\mu}_{i,t}$. We eliminate
action $i$ at the end of phase $t$, i.e., delete it from $A_s$, if
$\hat{\mu}_t^*-\hat{\mu}_{i,t} > \log(T/\delta)/\sqrt{t}$.
%So far identical to \SuccesiveElimination, the difference is how we
%continue after elimination.
In \SuccesiveEliminationReset we simply reset the algorithm with
$A=A_s-A_{e,t}$, where $A_{e,t}$ is the set of eliminated actions
after phase $t$. Namely, we restart $\hat{\mu}_{i,t}$ and ignore the
old rewards before the elimination.

\begin{lemma}
Consider algorithm \SuccesiveEliminationReset.
\begin{itemize}
\item[(a)]
With probability $1-\delta$, $\IReg(n)=O(\log(T/\delta)/\sqrt{n/K})$.

\item[(b)]
Assuming Bernoulli rewards, the algorithm is monotone.
\ascomment{Steven: for each \MRV, or only in expectation?}
\ascomment{Steven: can you derive optimal regret?}
\end{itemize}
\end{lemma}

\begin{proof}
\xhdr{Part (a).}
Let the best action be $a^*=\arg\max_a \mu_a$. With probability
$1-\delta$ at any time $n$ we have that for any action $i\in A_s$
that $|\hat{\mu}_i -\mu_i|\leq \log(T/\delta)/\sqrt{n/K}$, and
$a^*\in A_s$. This implies that any action $a$ such
$\mu_{a^*}-\mu_{a}> 3\log(T/\delta)/\sqrt{n/K}$ is eliminated.
Therefore, any action in $A_s$ has \BIR$(n)$ of at most
$6\log(T/\delta)/\sqrt{n/K}$.

\xhdr{Part (b).}
Consider the first time $T$ an action is eliminated, and let
$T=\tau$ be a realized value of $T$. Then, clearly for $n<\tau$ we
have $\rew(n)=\rew(1)$ .
